Gradual Soundness: Lessons from Static Python

### 2022-01-13 pj-7.1/paper.tex
pj: sec 4 pass

\newcommand{\spx}{\code{x}}
\newcommand{\spf}{\code{f}}
\newcommand{\spc}{\code{c}}
\newcommand{\spprog}{\codefont{prog}}
\newcommand{\spstmt}{\codefont{stmt}}
\newcommand{\spexpr}{\codefont{expr}}
\newcommand{\spvalue}{\codefont{val}}
\newcommand{\vdashsub}[1]{\vdash_{#1}}
\newcommand{\wtprog}{\vdashsub{\mathbf{P}}}
\newcommand{\wtexpr}{\vdashsub{\mathbf{E}}}
\newcommand{\spsubt}{\mathrel{<:}}
\newcommand{\spsubteq}{\mathrel{\leq:}}
\newcommand{\spcompat}{\sqsubseteq}
\newcommand{\spconsist}{\spcompat}
\newcommand{\spmatr}{\prec}
\newcommand{\spvardef}[3]{\sptvardef{#1}{#2} = #3}
\newcommand{\spfundef}[4]{\mathrm{def}~\spapp{#1}{#2}\code{ -> }#3: #4}
\newcommand{\spclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\spself}{\code{self}}
\newcommand{\spobject}{\code{object}}
\newcommand{\spnone}{\code{none}}
\newcommand{\spint}{\codefont{int}}
\newcommand{\spbool}{\codefont{bool}}
\newcommand{\spfloat}{\codefont{float}}
\newcommand{\sppydict}[1]{\code{\{}#1\code{\}}}
\newcommand{\spchkdict}[3]{\spapp{\paramtype{\code{chkdict}}{#1, #2}}{#3}}
\newcommand{\spenvapp}[2]{\spapp{#1}{#2}}
\newcommand{\spdictref}[2]{#1[#2]}
\newcommand{\spdictset}[3]{\spdictref{#1}{#2} = #3}
\newcommand{\spobjref}[2]{#1.#2}
\newcommand{\spobjset}[3]{\spobjref{#1}{#2} = #3}
\newcommand{\spobjapp}[3]{\spobjref{#1}{\spapp{#2}{#3}}}

@@

\newcommand{\mfapply}[2]{#1\,(#2)}
\newcommand{\mffont}[1]{\mathit{#1}}
\newcommand{\mftypeF}[1]{\mfapply{\mffont{F}}{#1}}
\newcommand{\mfcast}[2]{\mfapply{\mffont{Cast}}{#1, #2}}
\newcommand{\mfopt}[1]{\mfapply{\mffont{opt}}{#1}}

@@

\section{Model}
\label{s:model}
%% purpose = types, checks, and optimizations can be formalized and analyzed

%% 2022-01-06 TODO subtyping for classes C ... matters for casts at least ... just add a Gamma, follow chain up one step at a time until refl.
%% 2022-01-04 KC will look into pbt for nom properties (I don't think there's much to see there!)

%% NOTE
%% - remark: model gives errors of the same kind in SP, but not always the same
%%   message, after all its not our goal to reproduce SP exactly

@@

To validate the design of \SP{}, we have developed a mechanized model using PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model encompasses two soundness claims:
%% addresses two soundness concerns?
\begin{enumerate}
  \item
    \SP{}'s overall approach to gradual interactions is sound; namely, that interactions
    with untyped code do not undermine the validity of types.
  \item
    The typed half of \SP{} is a sound type system for Python.
\end{enumerate}
This section explores the first claim in depth by way of a tiny formal model
that illustrates the \SP{} gradual interaction strategy.
%% Compared to other gradual languages, these boundaries are quite restrictive---which
%% partly explains how it is able to perform well in practice~(\cref{s:eval}).
%% Indeed the boundaries are effectively a subset of those in Nom~\cite{mt-oopsla-2021}.
The second claim is justified by our Redex development;
in particular, by its test suite.
This section describes the Redex development in broad strokes.

\subsection{Surface Syntax and Types}
%% TODO assume well-formed; no circular defs (x:t = x)

\begin{figure}[t]
  \begin{langarray}
    \spprog & \defeq &
      \spstmt \langmid \spstmt,~\spprog
  \\
    \spstmt & \defeq &
      \spvardef{\spx}{\sptype}{\spexpr} \langmid
  \\ & &
      \spfundef{\spf}{\spann{\spx}{\sptype}}{\sptype}{\spexpr} \langmid
  \\ & &
      \spclassdef{\spc}{\spc}{\spvardef{\spx}{\sptype}{\spexpr}}{\spfundef{\spf}{\spself, \spann{\spx}{\sptype}}{\sptype}{\spexpr}}
  \\
    \spexpr & \defeq &
      \spx \langmid \spnone \langmid \spint \langmid \spbool \langmid \spfloat \langmid \spobject \langmid
      \sppydict{\spx: \spexpr, \ldots} \langmid
      \spchkdict{\sptype}{\sptype}{\sppydict{\spx: \spexpr, \ldots}} \langmid
  \\ & &
      \spapp{\spx}{\spexpr} \langmid
      \spdictref{\spexpr}{\spexpr} \langmid
      \spdictset{\spexpr}{\spexpr}{\spexpr} \langmid
      \spobjref{\spc}{\spx} \langmid
      \spobjset{\spc}{\spx}{\spexpr} \langmid
      \spobjapp{\spc}{\spf}{\spexpr}
  \\
    \sptype & \defeq &
      \sptdyn \langmid
      \sptnone \langmid
      \sptint \langmid
      \sptbool \langmid
      \sptfloat \langmid
      \sptobject \langmid
      \sptclass \langmid
  \\ & &
      \sptpydict{\sptype}{\sptype} \langmid
      \sptchkdict{\sptype}{\sptype} \langmid
      \sptunion{\sptype, \ldots}
  \\
    \sptenv & \defeq &
      \sptenvnil \langmid
      \sptvardef{\spx}{\sptype},~\sptenv \langmid
      \sptfundef{\spf}{\sptype}{\sptype},~\sptenv \langmid
      \sptclassdef{\spc}{\spc}{\sptvardef{\spx}{\sptype}}{\sptfundef{\spf}{\sptclass, \sptype}{\sptype}},~\sptenv
  \\
    \spx, \spf, \spc & \defeq & \mbox{variable names}
  \end{langarray}

@@

The \SP{} surface syntax is the same as Python 3.8.
Programmers may write type annotations using the PEP 484 syntax;
unannotated code gets the dynamic type by default.
The available static types provide a nominal type system with first-order functions.
These surface types are meant to be descriptive.
By design, they are not fully sound.
For this reason, we omit the surface typing judgment and merely present their
declarative syntax.
%% TODO awkward

\Cref{f:surface-types} illustrates the key aspects of the \SP{} syntax and types
using a tiny grammar.
A program is a sequence of statements;
a statement defines a variable, function, or class.
These definitions may only appear on the top level
and they all require full type annotations.
Functions must have one positional argument.
Classes must declare one parent (perhaps $\spobject$), one field, and one method.
Expressions describe values and simple computations.
The basic values consist of the following: the none value, integers,
booleans (which are the integers \code{0} and \code{1}), floating point numbers, and the top object.
There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
The remaining forms express function calls, dictionary reads and writes, object field reads and writes,
and method calls.
Types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
user-defined class, and union types.
We assume that all unions are written a concise normal form, e.g., that
$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be flattened and normalized to $\sptdyn$.
There is no function type; the only way to define a higher-order function is to use the dynamic type.

%% FILL generally, no way to express higher-order cast ... either trivial C=C or impossible C=D

Relative to \SP{} and our Redex model, the formalization in \cref{f:surface-types} omits
many details of Python including class variables, imports, conditionals, and exception handlers.
These details are crucial in the context of the Redex model, which tests whether \SP{}
soundly approximates Python.
They are less important here, where our focus is on type boundaries.
None of the omitted features give substantially new ways for typed code to interact with dynamic code.

@@

  \bigskip
  \(
    \mftypeF{\sptype_0}
    \mfeq
    \left\{\begin{array}{ll}
      \sptrawpydict & \mbox{if $\sptype_0 = \sptpydict{\sptype}{\sptype}$}
    \\
      \sptchkdict{\mftypeF{\sptype_1}}{\mftypeF{\sptype_2}} & \mbox{if $\sptype_0 = \sptchkdict{\sptype_1}{\sptype_2}$}
    \\
      \sptoptional{\mftypeF{\sptype_1}} & \mbox{if $\sptype_0 = \sptoptional{\sptype_1}$}
    \\
      \sptdyn & \mbox{if $\sptype_0 = \sptunion{\sptype, \ldots}$ and $\sptype_0 \neq \sptoptional{\sptype}$}
    \\
      \sptype_0 & \mbox{otherwise}
    \end{array}\right.
  \)

@@

Evaluation type $\spteval$ are a subset of the surface types.
These are the types that \SP{} promises to soundly enforce.
\Cref{f:surface-to-eval-types} presents the syntax of evaluation types
and a retraction $\mftypeF{\cdot}$ from surface types to evaluation types.
As the retraction makes clear, the evaluation types make two main changes:
\begin{enumerate}
  \item
    The parameterized type for Python dictionaries
    $\sptpydict{\sptype_0}{\sptype_1}$ gets replaced with
    a raw type $\sptrawpydict$.
    The raw type behaves the same as $\sptpydict{\sptdyn}{\sptdyn}$ would.
  \item
    Unions get replaced with the dynamic type, except for the special case
    of unions with none ($\sptoptional{\sptype_0}$).
\end{enumerate}

The \SP{} team chose these particular evaluation types because they can all
be enforced fully and efficiently with run-time casts.

@@

\Cref{t:cast} describes the casts in detail.
Most call for a simple tag check, i.e., a Python \code{isinstance} test.
The sole exception is optional types, which require a tag check and a test for the none value.
Even checked dictionaries rely on tag checks.
A checked dict type with keys $\sptype_0$ and values $\sptype_1$ accepts only
checked dict values that were initialized with exactly the same key and value type.
As noted in~\cref{s:type-dynamic}, the exact-match rules applies even when $\sptype_0$
or $\sptype_1$ is the dynamic type.

There are two key observations to make about casts.
First, all allowed casts can run in constant time.
No cast requires traversing a data structure.
Similarly, no cast allocates a wrapper to check higher-order behaviors in a delayed fashion.
This latter property means that blame-tracking is trivial;
casts succeed or fail immediately.

Second, the strict cast for checked dictionaries makes type compatibility simple in \SP{}.
Values that are partly-dynamic are incompatible with static types.
For example, suppose that a function expects a list of numbers.
In a typical gradual language, a list of dynamic values would be a valid input to the function.
In \SP{}, a list of dynamic values is not a valid input.
Consequently, \SP{} can use a simple ahead-of-time strategy to enforce soundness.
Any position where the dynamic type meets a static type needs a cast.
Other meetings between types can be resolved statically.

@@

To better show where \SP{} needs to insert casts, we present a selection of typing rules.
Recall that a program declares variables, functions, and classes~(\cref{f:surface-types}).
These statements come with types and fill an environment ($\sptenv$).
Relative to the current environment, the expressions within each statement must
satisfy the judgment in~\cref{f:eval-types}

The first two rules are for function application.
A typed function may be applied an argument that matches its domain type,
in which case it computes a value that matches its codomain type.
A dynamically-typed variable may be applied to any input (that matches the top
type $\sptobject$) to yield a dynamically-typed result.
The next rules illustrate writes to Python dictionaries and checked dictionaries.
A Python dict may be updated with any kind of key and value.
By contrast, a checked dict requires keys and values that match its type parameters.
Next we have two rules for classes:
\trule{F-Set} says that writes to a class field must match the declared field
type; and \trule{M-App} shows that typed methods impose similar constraints as
typed fuctions.

The final two rules, \trule{C-Sub} and \trule{Matr}, depend on auxiliary judgments
for consistent subtyping ($\spcompat$) and materialization ($\spmatr$).
Consistent subtyping relates type $\spteval_0$ to type $\spteval_1$ if they are
related by static subtyping ($\spteval_0 \spsubteq \spteval_1$) or if
$\spteval_1$ is the dynamic type.
When this relation holds, it is safe to upcast a value from type $\spteval_0$
to type $\spteval_1$.
Materialization relates the dynamic type to any non-dynamic type.
Occurrences of the \trule{Matr} rule are downcasts that require a
run-time check.

The name \emph{materialization} and the rule \trule{Matr} are inspired by prior
work~\cite{clps-popl-2019}.
By itself, the judgment is merely an upside-down type precision relation~\cite{g-icfp-2013,sv-dls-2008}.
Combined with the typing rule, however, materialization is a concise way to find
where a well-typed program needs to insert casts to ensure soundness.
For example, suppose that $\spf$ is a function from integers to the dynamic
type and that $\spx$ is a variable with the dynamic type.
An application $\spapp{\spf}{\spx}$ can satisfy type $\sptfloat$ using two materializations:

@@

We end here, with materialization, rather than present a semantics
for the formalization.
After all, the main benefit of a full formal semantics is to validate
the behavior of complex expressions---a job
better left to the mechanized Redex model.

Furthermore, our casts-via-materialization rule is more aggressive
than what \SP{} actually does to insert casts because
\SP{} supports interactions with plain Python code.
Because it chooses not to recompile Python modules, \SP{} inserts
casts in an overapproximate way:
\begin{itemize}
  \item
    typed functions and methods begin by checking all their arguments;
  \item
    typed classes check all field writes;
  \item
    checked dictionaries check all writes; and
  \item
    typed code checks the results of function calls, references, etc. whenever there is a materialization from the dynamic type.
\end{itemize}
\Cref{s:impl} explains
how the type-directed optimizer avoids some
of these checks.

\begin{figure}[t]
  \fbox{$\sptenv \wtexpr \spexpr : \spteval$}~selected rules
  \begin{mathpar}
    \inferrule*[left=S-App]{
      \sptfundef{\spf_0}{\spteval_0}{\spteval_1} \in \sptenv_0
      \\\\
      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
    }{
      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_0} : \spteval_1
    }

@@

    \inferrule*[left=C-Sub]{
      \sptenv_0 \wtexpr \spexpr_0 : \spteval_1
      \\
      \spteval_1 \spcompat \spteval_0
    }{
      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
    }

    \inferrule*[left=Matr]{
      \sptenv_0 \wtexpr \spexpr_0 : \spteval_1
      \\
      \spteval_1 \spmatr \spteval_0
      %% side condition: result type is not a primitive
    }{
      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
    }

@@

  \begin{minipage}[t]{0.5\columnwidth}
    \fbox{$\spteval \spcompat \spteval$}
    \begin{mathpar}
      \inferrule*{
        \spteval_0 \spsubt \spteval_1
      }{
        \spteval_0 \spcompat \spteval_1
      }

      \inferrule*{
      }{
        \spteval_0 \spcompat \sptdyn
      }
  \end{mathpar}

  \end{minipage}\begin{minipage}[t]{0.5\columnwidth}
    \fbox{$\spteval \spmatr \spteval$}
    \begin{mathpar}
      \inferrule*[right={$\spteval_0 \neq \sptdyn$}]{
      }{
        \sptdyn \spmatr \spteval_0
      }
    \end{mathpar}
  \end{minipage}

  \medskip
  \fbox{$\spteval \spsubteq \spteval$}~reflexive, transitive closure of the following $\spsubt$ relation:
  \begin{mathpar}
    \inferrule*{
    }{
      \spteval_0 \spsubt \sptobject
    }

    \inferrule*{
    }{
      \sptbool \spsubt \sptint
    }

    \inferrule*{
      \spteval_0 \spsubt \spteval_1
    }{
      \spteval_0 \spsubt \sptoptional{\spteval_1}
    }

@@

The central question for the model is whether \SP{}'s approach to gradual typing
is sound.
No matter how an expression mixes typed and untyped code, its static
\emph{evaluation} type should be an accurate prediction about the results
of evaluation.
Using the metavariable $\spvalue$ to represent heap-allocated Python values
and $\sprred$ as the evaluation relation (which would need to maintain a heap),
an approximate soundness theorem is as follows:

\begin{theorem}[Type Soundness]
  If\ \(~\vdash \spexpr : \sptype\)
  then one of the following holds:
  \begin{itemize}
    \item
      \(\spexpr \sprred \spvalue
        \mbox{ and }
        \vdash \spvalue : \mftypeF{\sptype}
      \)
    \item
      \(\spexpr\) diverges
    \item
      \(\spexpr\) reduces to an allowed error
  \end{itemize}
\end{theorem}

As mentioned above, our argument for soundness has two parts
and the question at hand is whether the boundaries to type dynamic
are adequately protected with run-time checks.
The Redex model's comprehensive test suite provides the other half of the
argument; namely, that \SP{} is a sound type system for Python.

@@

Alternatively, observe that the boundaries in \SP{} are a subset
of the boundaries in Nom, a gradual language that comes with a detailed
proof of soundness~\cite{mt-oopsla-2017,mt-oopsla-2021}.
Given an \SP{} program, it is possible to derive a Nom program with
the same type boundaries by replacing every checked dict instance
with a fresh class type.
Because different checked dict instantiations are incompatible with
one another, the \SP{} boundaries impose the same constraints as the Nom
boundaries.

\subsection{Redex Development, \SP{} Conformance}
%% pbt for soundness?

>>>>

\newcommand{\spx}{\code{x}}
\newcommand{\spf}{\code{f}}
\newcommand{\spc}{\sptclass}
\newcommand{\spprog}{\codefont{prog}}
\newcommand{\spstmt}{\codefont{stmt}}
\newcommand{\spexpr}{\codefont{expr}}
\newcommand{\spvalue}{\codefont{val}}
\newcommand{\vdashsub}[1]{\vdash_{#1}}
\newcommand{\wtprog}{\vdashsub{\mathbf{P}}}
\newcommand{\wtexpr}{\vdashsub{\mathbf{E}}}
\newcommand{\wtsub}{\vdash}
\newcommand{\spsubt}{\mathrel{<:}}
\newcommand{\spsubteq}{\mathrel{\leq:}}
\newcommand{\spcompat}{\sqsubseteq}
\newcommand{\spconsist}{\spcompat}
\newcommand{\spmatr}{\prec}
\newcommand{\spvardef}[3]{\sptvardef{#1}{#2} = #3}
\newcommand{\spfundef}[4]{\mathrm{def}~\spapp{#1}{#2}\code{ -> }#3: #4}
\newcommand{\spclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\spself}{\code{self}}
\newcommand{\spobject}{\code{object}}
\newcommand{\spnone}{\code{none}}
\newcommand{\spint}{\codefont{int}}
\newcommand{\spbool}{\codefont{bool}}
\newcommand{\spfloat}{\codefont{float}}
\newcommand{\sppydict}[1]{\code{\{}#1\code{\}}}
\newcommand{\spchkdict}[3]{\spapp{\paramtype{\code{chkdict}}{#1, #2}}{#3}}
\newcommand{\spenvapp}[2]{\spapp{#1}{#2}}
\newcommand{\spdictref}[2]{#1[#2]}
\newcommand{\spdictset}[3]{\spdictref{#1}{#2} = #3}
\newcommand{\spobjref}[2]{#1.#2}
\newcommand{\spobjset}[3]{\spobjref{#1}{#2} = #3}
\newcommand{\spobjapp}[3]{\spobjref{#1}{\spapp{#2}{#3}}}

@@

\newcommand{\mfapply}[2]{#1\,(#2)}
\newcommand{\mffont}[1]{\mathit{#1}}
\newcommand{\mftypeF}[1]{\mfapply{\mffont{F}}{#1}} % {\mfapply{\mffont{T}_{\downarrow}\!\mffont{S}}{#1}}
\newcommand{\mfcast}[2]{\mfapply{\mffont{Cast}}{#1, #2}}
\newcommand{\mfopt}[1]{\mfapply{\mffont{opt}}{#1}}

@@

\section{Model}
\label{s:model}

@@

To validate the design of \SP{}, we have developed a model of the language in PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model encompasses two soundness claims:
%% addresses two soundness concerns?
\begin{enumerate}
  \item
    \SP{}'s overall approach to gradual interactions is sound; in particular,
    interactions with untyped code do not undermine the validity of types.
  \item
    The typed half of \SP{} is a sound type system for Python.
\end{enumerate}
This section explores the first claim in depth by way of a small formalization
(as opposed to the Redex \emph{mechanization})
that illustrates the \SP{} gradual interaction strategy.
%% Compared to other gradual languages, these boundaries are quite restrictive---which
%% partly explains how it is able to perform well in practice~(\cref{s:eval}).
%% Indeed the boundaries are effectively a subset of those in Nom~\cite{mt-oopsla-2021}.
The second claim is justified by our Redex development and its test suite~(\cref{s:redex}).

\subsection{Surface Syntax and Types}

\begin{figure}[t]
  \begin{langarray}
    \spprog & \defeq &
      \spstmt \langmid \spstmt,~\spprog
  \\
    \spstmt & \defeq &
      \spvardef{\spx}{\sptype}{\spexpr} \langmid
  \\ & &
      \spfundef{\spf}{\spann{\spx}{\sptype}}{\sptype}{\spexpr} \langmid
  \\ & &
      \spclassdef{\spc}{\spc}{\spvardef{\spx}{\sptype}{\spexpr}}{\spfundef{\spf}{\spself, \spann{\spx}{\sptype}}{\sptype}{\spexpr}}
  \\
    \spexpr & \defeq &
      \spx \langmid \spnone \langmid \spint \langmid \spbool \langmid \spfloat \langmid \spobject \langmid
      \sppydict{\spx: \spexpr, \ldots} \langmid
      \spchkdict{\sptype}{\sptype}{\sppydict{\spx: \spexpr, \ldots}} \langmid
  \\ & &
      \spapp{\spx}{\spexpr} \langmid
      \spdictref{\spexpr}{\spexpr} \langmid
      \spdictset{\spexpr}{\spexpr}{\spexpr} \langmid
      \spobjref{\spc}{\spx} \langmid
      \spobjset{\spc}{\spx}{\spexpr} \langmid
      \spobjapp{\spc}{\spf}{\spexpr}
  \\
    \sptype & \defeq &
      \sptdyn \langmid
      \sptnone \langmid
      \sptint \langmid
      \sptbool \langmid
      \sptfloat \langmid
      \sptobject \langmid
      \sptclass \langmid
  \\ & &
      \sptpydict{\sptype}{\sptype} \langmid
      \sptchkdict{\sptype}{\sptype} \langmid
      \sptunion{\sptype, \ldots}
  \\
    \sptenv & \defeq &
      \sptenvnil \langmid
      \sptvardef{\spx}{\sptype},~\sptenv \langmid
      \sptfundef{\spf}{\sptype}{\sptype},~\sptenv \langmid
      \sptclassdef{\spc}{\spc}{\sptvardef{\spx}{\sptype}}{\sptfundef{\spf}{\sptclass, \sptype}{\sptype}},~\sptenv
  \\
    \spx, \spf & \defeq & \mbox{variable names}
  \end{langarray}

@@

\Cref{f:surface-types} illustrates key aspects of the \SP{} syntax and types.
A program is a sequence of statements;
a statement defines a variable, function, or class.
These definitions may only appear on the top level
and they all require full type annotations.
Functions must have one positional argument.
Classes must declare one parent (perhaps $\spobject$), one field, and one method.
Expressions describe values and simple computations.
The basic values consist of the following: the none value, integers,
booleans (which are the integers \code{0} and \code{1}), floating point numbers, and the top object.\footnote{By contrast to
Python, in which \code{int} is a type and \code{None} is a value, our formalization uses lowercase names for
values and capitalized names for types.}
There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
The remaining forms express function calls, dictionary reads and writes, object field reads and writes,
and method calls.
Types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
user-defined class, and union types.
We assume that all unions are written in a tidy normal form, e.g., that
$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be flattened and normalized to $\sptdyn$.
There is no function type in the model, or in \SP{} for that matter.
The only way to define a higher-order function is to use the dynamic type.

The surface types are meant to describe programs, but they are not designed
to be fully sound.
For this reason, we omit the surface typing judgment, which scans for
logical errors, and merely present the declarative syntax of surface types.
We also omit a well-formedness judgment that checks for circular definitions.

Relative to \SP{} and our Redex mechanization, the formalization in \cref{f:surface-types} omits
many details of Python including class variables, imports, conditionals, and exception handlers.
These details are crucial in the mechanization, which tests whether \SP{}
soundly approximates Python.
They are less important here, where our focus is on type boundaries.
None of the omitted features give substantially new ways for typed code to interact with dynamic code.

@@

  \bigskip
  \(
    \mftypeF{\sptype_0}
    \mfeq
    \left\{\begin{array}{ll}
      \sptrawpydict & \mbox{if $\sptype_0 = \sptpydict{\sptype}{\sptype}$}
    \\
      \sptoptional{\mftypeF{\sptype_1}} & \mbox{if $\sptype_0 = \sptoptional{\sptype_1}$}
    \\
      \sptdyn & \mbox{if $\sptype_0 = \sptunion{\sptype, \ldots}$ and $\sptype_0 \neq \sptoptional{\sptype}$}
    \\
      \sptchkdict{\mftypeF{\sptype_1}}{\mftypeF{\sptype_2}} & \mbox{if $\sptype_0 = \sptchkdict{\sptype_1}{\sptype_2}$}
    \\
      \sptype_0 & \mbox{otherwise}
    \end{array}\right.
  \)

@@

Evaluation types $\spteval$ are a subset of the surface types.
These are the types that \SP{} promises to soundly enforce.
\Cref{f:surface-to-eval-types} presents the syntax of evaluation types
and a retraction $\mftypeF{\cdot}$ from surface types to evaluation types.
As the retraction shows, the evaluation types make two main changes:
\begin{enumerate}
  \item
    The parameterized type for Python dictionaries
    $\sptpydict{\sptype_0}{\sptype_1}$ gets replaced with
    a raw type $\sptrawpydict$.
    The raw type behaves the same as $\sptpydict{\sptdyn}{\sptdyn}$ would.
  \item
    Unions get replaced with the dynamic type, except for the special case
    of unions with none ($\sptoptional{\sptype_0}$).
\end{enumerate}

@@

\SP{} evaluation types can all be enforced fully and in constant-time
with casts.
\Cref{t:cast} describes the casts in detail.
Most call for a tag check, i.e., a Python \code{isinstance} test.
The sole exception is optional types, which require a tag check and a test for the none value.
Even checked dictionaries rely on tag checks.
A checked dict type with keys $\spteval_0$ and values $\spteval_1$ accepts only
checked dict values that were initialized with exactly the same key and value type.
As noted in~\cref{s:type-dynamic}, the exact-match rules applies even when $\spteval_0$
or $\spteval_1$ is the dynamic type.

As promised, all casts run in constant time.
No cast requires traversing a data structure.
Similarly, no cast allocates a wrapper to check higher-order behaviors in a delayed fashion.
This latter property means that blame-tracking is trivial because
casts either succeed or fail immediately.
%% In other words, \SP{} provides \emph{immediate
%% accountability}~\cite{mt-oopsla-2017} for cast errors.

The strict cast for checked dictionaries greatly simplifies type compatibility for \SP{}.
Values that are partly-dynamic are simply incompatible with static types.
For example, suppose that a function expects a list of numbers.
In a typical gradual language, a list of dynamic values would be a valid input to the function.
In \SP{}, a list of dynamic values is not a valid input.
Consequently, \SP{} can use an ahead-of-time strategy to enforce soundness.
Any position where the dynamic type meets a static type needs a cast.
Other meetings between types can be resolved statically.
\SP{} does not need to worry that a list of dynamic values might be masquerading
under a fully-static type.

@@

To better show where \SP{} needs to insert casts, we present a selection of typing rules.
Recall that a program declares variables, functions, and classes~(\cref{f:surface-types}).
These statements come with types and fill an environment ($\sptenv$).
Relative to the current environment, the expressions within each statement must
satisfy the typing judgment in~\cref{f:eval-types}

The first two rules are for function application.
A typed function may be applied to an argument that matches its domain type,
in which case it computes a value that matches its codomain type.
A dynamically-typed variable may be applied to any input (that matches the top
type $\sptobject$) to yield a dynamically-typed result.
The next rules illustrate writes to Python dictionaries and checked dictionaries.
A shallow Python dict may be updated with any kind of key and value.
By contrast, a concrete checked dict requires keys and values that match its type parameters.
Next we have two rules for classes:
\trule{F-Set} says that writes to a class field must match the declared field
type; and \trule{M-App} shows that typed methods impose similar constraints as
typed fuctions.

The final two rules, \trule{C-Sub} and \trule{Matr}, depend on auxiliary judgments
for consistent subtyping ($\spcompat$) and materialization ($\spmatr$).
Consistent subtyping relates type $\spteval_0$ to type $\spteval_1$ if they are
related by static subtyping ($\spsubteq$) or if
$\spteval_1$ is the dynamic type.
When this relation holds, it is safe to upcast a value from type $\spteval_0$
to type $\spteval_1$.
Materialization relates the dynamic type to any non-dynamic type.
Occurrences of the \trule{Matr} rule are downcasts that require a
run-time check.

The name \emph{materialization} and the rule \trule{Matr} are inspired by prior
work~\cite{clps-popl-2019}.
By itself, the judgment is merely an upside-down type precision relation~\cite{g-icfp-2013,sv-dls-2008}.
Combined with the typing rule, however, materialization is a concise way to find
where a well-typed program needs casts to ensure soundness.
For example, suppose that $\spf$ is a function from integers to the dynamic
type and that $\spx$ is a variable with the dynamic type.
An application $\spapp{\spf}{\spx}$ can satisfy the type $\sptfloat$ using two materializations:

@@

We end here, with materialization, rather than present a semantics
for the formalization.
After all, the main benefit of a full formal semantics is to validate
the behavior of complex expressions---and this job is
better left to the Redex mechanization.

Furthermore, our casts-via-materialization rule is more aggressive
than what \SP{} actually does to insert casts.
Because \SP{} supports interactions with unanalyzed Python modules, it inserts
casts in an overapproximate way:
\begin{itemize}
  \item
    typed functions and methods begin by checking all their arguments;
  \item
    typed classes check all field writes;
  \item
    checked dictionaries check all writes; and
  \item
    typed code checks the results of function calls, references, and other elimination forms
    whenever there is a materialization from the dynamic type.
\end{itemize}
\Cref{s:impl} explains
how the type-directed optimizer helps avoid some
of these checks.

\begin{figure}[p]
  \fbox{$\sptenv \wtexpr \spexpr : \spteval$}~selected rules
  \begin{mathpar}
    \inferrule*[left=S-App]{
      \sptfundef{\spf_0}{\spteval_0}{\spteval_1} \in \sptenv_0
      \\\\
      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
    }{
      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_0} : \spteval_1
    }

@@

    \inferrule*[left=C-Sub]{
      \sptenv_0 \wtexpr \spexpr_0 : \spteval_1
      \\\\
      \sptenv_0 \wtsub \spteval_1 \spcompat \spteval_0
    }{
      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
    }

    \inferrule*[left=Matr]{
      \sptenv_0 \wtexpr \spexpr_0 : \spteval_1
      \\\\
      \spteval_1 \spmatr \spteval_0
      %% side condition: result type is not a primitive
    }{
      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
    }

@@

  \begin{minipage}[t]{0.6\columnwidth}
    \fbox{$\sptenv \wtsub \spteval \spcompat \spteval$}
    \begin{mathpar}
      \inferrule*{
        \sptenv_0 \wtsub \spteval_0 \spsubt \spteval_1
      }{
        \sptenv_0 \wtsub \spteval_0 \spcompat \spteval_1
      }

      \inferrule*{
      }{
        \sptenv_0 \wtsub \spteval_0 \spcompat \sptdyn
      }
  \end{mathpar}

  \end{minipage}\begin{minipage}[t]{0.4\columnwidth}
    \fbox{$\spteval \spmatr \spteval$}
    \begin{mathpar}
      \inferrule*[right={$\spteval_0 \neq \sptdyn$}]{
      }{
        \sptdyn \spmatr \spteval_0
      }
    \end{mathpar}
  \end{minipage}

  \medskip
  \fbox{$\sptenv \wtsub \spteval \spsubteq \spteval$}~reflexive, transitive closure of the following $\spsubt$ relation:
  \begin{mathpar}
    \inferrule*{
    }{
      \sptenv_0 \wtsub \spteval_0 \spsubt \sptobject
    }

    \inferrule*{
    }{
      \sptenv_0 \wtsub \sptbool \spsubt \sptint
    }

    \inferrule*{
      \sptenv_0 \wtsub \spteval_0 \spsubt \spteval_1
    }{
      \sptenv_0 \wtsub \spteval_0 \spsubt \sptoptional{\spteval_1}
    }

    \inferrule*{
      \sptclassdef{\sptclass_0}{\sptclass_2}{\ldots}{\ldots} \in \sptenv_0
      \\
      \sptenv_0 \wtsub \sptclass_2 \spsubt \sptclass_1
    }{
      \sptenv_0 \wtsub \sptclass_0 \spsubt \sptclass_1
    }

@@

The central question for the model is whether \SP{}'s approach to gradual typing
is sound.
No matter how an expression mixes typed and untyped code, its static
\emph{evaluation type} should be an accurate prediction about the results
of evaluation.
Using the metavariable $\spvalue$ to represent heap-allocated Python values
and $\sprred$ as the evaluation relation (which would need to maintain a heap),
an approximate soundness theorem is as follows:

\begin{theorem}[Type Soundness]
  If\ \(~\vdash \spexpr : \sptype\)
  then one of the following holds:
  \begin{itemize}
    \item
      \(\spexpr \sprred \spvalue
        \mbox{ and }
        \vdash \spvalue : \mftypeF{\sptype}
      \)
    \item
      \(\spexpr\) reduces to an allowed error
    \item
      \(\spexpr\) diverges
  \end{itemize}
\end{theorem}

As mentioned above, our argument for soundness has two parts
The question at hand is whether the boundaries to type dynamic
are adequately protected with run-time checks.
The Redex mechanization's comprehensive test suite provides the other half of the
argument; namely, that \SP{} is a sound type system for Python.

@@

Alternatively, observe that the boundaries in \SP{} are a subset
of the possible boundaries in Nom, a gradual language that comes with a detailed
proof of soundness~\cite{mt-oopsla-2017,mt-oopsla-2021}.
Given an \SP{} program, we can derive a Nom program with
the same type boundaries by replacing every checked dict instance
with a fresh class type.
Because distinct checked dict instantiations are incompatible,
the \SP{} boundaries impose the same constraints as the translated Nom
boundaries.

\subsection{Redex Development, \SP{} Conformance}
\label{s:redex}
%% pbt for soundness?
